Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on

### 2011-06-03 paper.tex
Finish figures

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse is an award-winning
molecular workbench for life sciences
that wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
preparing and analysing computational chemistry calculations. Both
projects are designed to be extended or scripted by users through
the provision of a plugin architecture and scripting support (using
Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
of Avogadro). An interesting aspect of both Avogadro and Bioclipse is
that they share some developers with the underlying toolkits and this
has driven the development of new features in the CDK and Open Babel.

@@

In 2009, iChemLabs released the ChemDoodle Web Components
library~\cite{ChemDoodleWeb} under the GPL v3 license (with a
liberal HTML exception). This library is completely implemented in JavaScript
and uses HTML5 to allow the scientist
to present publication quality 2D and 3D graphics and animations for
chemical structures, reactions and spectra. Beyond graphics, this tool
provides a framework for user interaction to create dynamic
applications through web browsers, desktop platforms and mobile
devices such as the iPhone, iPad and Android devices.

@@

One of the successes of the Blue Obelisk has been to bring developers
together from different Open Source chemistry projects so that they
look for opportunities to collaborate rather than compete, and to
leverage work done by other projects to avoid duplication of effort.
As an example of this, when in March 2008 the Jmol development team
were looking to add support for energy minimisation, rather than
implement a forcefield from scratch they ported the UFF forcefield~\cite{Rappe:1992um}
implementation from Open Babel to Jmol. This code enables Jmol to
support 2D to 3D conversion of structures (through energy
minimisation). In a similar manner, efficient Jmol code for atom-atom rebonding
has been ported to the CDK. Figure 2 shows the collaborative nature of
software developed in the Blue Obelisk, as one project builds on
functionality provided by another project.

@@

While each Blue Obelisk project has its own website and point of
contact (typically a mailing list), because of the breadth of BO
projects it can be difficult for a newcomer to understand which of
them, if any, can best address a particular problem. To address this
issue, members of the Blue Obelisk established a Question \& Answer
website\cite{WebBOShapado} (see Figure~3).  This is a website in the
style of Stack Overflow\cite{WebStackOverflow} that encourages high quality answers (and
questions) through the use of a voting system. In the year since it
was established, over 200 users have registered, many of whom had no
previous involvement with the Blue Obelisk, showing that the Q\&A
website complements earlier existing channels of communication.

@@

  \subsection*{Figure 1 - Dependency diagram of some Blue Obelisk projects.}
      Each block represents a project. Square blocks show Open Data, ovals are Open Source,
      and diamonds are Open Standards. %%Colors represent license: LGPL is green, GPL is orange,
      %%and BSD is blue.

  \subsection*{Figure 3 - Screenshot of the Blue Obelisk eXchange Question
    and Answer website.}

>>>>

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse (Figure~1) is an award-winning
molecular workbench for life sciences
that wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors while Avogadro (Figure~2) is a 3D molecular editor and viewer aimed at
preparing and analysing computational chemistry calculations. Both
projects are designed to be extended or scripted by users through
the provision of a plugin architecture and scripting support (using
Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
of Avogadro). An interesting aspect of both Avogadro and Bioclipse is
that they share some developers with the underlying toolkits and this
has driven the development of new features in the CDK and Open Babel.

@@

In 2009, iChemLabs released the ChemDoodle Web Components
library~\cite{ChemDoodleWeb} under the GPL v3 license (with a
liberal HTML exception). This library is completely implemented in JavaScript
and uses HTML5 to allow the scientist
to present publication quality 2D and 3D graphics (see Figure~3) and animations for
chemical structures, reactions and spectra. Beyond graphics, this tool
provides a framework for user interaction to create dynamic
applications through web browsers, desktop platforms and mobile
devices such as the iPhone, iPad and Android devices.

@@

One of the successes of the Blue Obelisk has been to bring developers
together from different Open Source chemistry projects so that they
look for opportunities to collaborate rather than compete, and to
leverage work done by other projects to avoid duplication of effort.
As an example of this, when in March 2008 the Jmol development team
were looking to add support for energy minimisation, rather than
implement a forcefield from scratch they ported the UFF forcefield~\cite{Rappe:1992um}
implementation from Open Babel to Jmol. This code enables Jmol to
support 2D to 3D conversion of structures (through energy
minimisation). In a similar manner, efficient Jmol code for atom-atom rebonding
has been ported to the CDK. Figure~4 shows the collaborative nature of
software developed in the Blue Obelisk, as one project builds on
functionality provided by another project.

@@

While each Blue Obelisk project has its own website and point of
contact (typically a mailing list), because of the breadth of BO
projects it can be difficult for a newcomer to understand which of
them, if any, can best address a particular problem. To address this
issue, members of the Blue Obelisk established a Question \& Answer
website\cite{WebBOShapado} (see Figure~5).  This is a website in the
style of Stack Overflow\cite{WebStackOverflow} that encourages high quality answers (and
questions) through the use of a voting system. In the year since it
was established, over 200 users have registered, many of whom had no
previous involvement with the Blue Obelisk, showing that the Q\&A
website complements earlier existing channels of communication.

@@

 \subsection*{Figure 1 - Screenshot of Bioclipse using Jmol to
visualise a molecular surface}

 \subsection*{Figure 2 - Screenshot of Avogadro showing a depiction of
a carbon nanotube}

 \subsection*{Figure 3 - Screenshot of the MolGrabber 3D demo from
ChemDoodle Web Components}

  \subsection*{Figure 4 - Dependency diagram of some Blue Obelisk projects.}
Each block represents a project. Square blocks show Open Data, ovals are Open Source,
and diamonds are Open Standards.

  \subsection*{Figure 5 - Screenshot of the Blue Obelisk eXchange Question
    and Answer website.}
